The Heavy Photon Search Test Detector

### 2014-04-03 hps-testrun-nim.tex
Comments by Sho.

The trigger system is designed to select time coincidences of electromagnetic clusters in the 
top and bottom halves of the ECal. Figure~\ref{fig:hps_trigger_cal} shows a schematic overview 
of each stage of the system. 
 \begin{figure}[b]
\begin{center}
{\small
 \includegraphics[width=8cm]{figures/hps_trigger_cal}
\caption{Block diagram of the ECAL trigger system consisting of the FADC that samples and digitizes 
signals for each detector channel and sends them for cluster finding in the CTP. The CTP clusters are 
sent to the SSP where the final trigger decision is taken based on pairs of clusters in both halves of the 
ECal. The decision is sent to the Trigger Supervisor (TS) that generates the necessary signals to readout 
the sub-detectors.}
 \label{fig:hps_trigger_cal}
}
\end{center}
 \end{figure}
Each channel on the FADC board have an 
independent data path to send 5-bit pulse energy and 3-bit pulse arrival time information every 32~ns 
to a 
trigger processing board (CTP) in the same crate every. Contrary to the readout path 
described in Sec.~\ref{sec:fadc}, this energy is a pedestal subtracted time-over-threshold sum with 
programmable offsets and minimum threshold discriminator for each channel. With input from all 
FADC channels, i.e. one half of the ECal, the CTP performs cluster finding and calculates cluster 
energy, shape and timing information. The 3x3 fixed-window, highly parallel, FPGA-based cluster 
algorithm simultaneously searches for up to 125 clusters with energy sum larger than 
the programmable energy threshold ($\approx270$~MeV). Crystals in the 
fixed-window are included in the sum if the leading edge of the pulse occurred within a 8~ns time 
window (4~ns minimum). To take into account clock skew and jitter throughout the system, the
clustering algorithm can allow for crystal hits from up to seven previous or future 4~ns clock cycles to 
contribute to the cluster sum.  A separate sub-system board (SSP) receives the clusters (only the 
highest energy 3x3 window to deal with overlapping and very large clusters) from 
the top and bottom half CTP every 32~ns and searches for pairs of clusters in a 8~ns wide 
trigger window. The trigger supervisor (TS) generates all the necessary signals and controls the 
entire DAQ system readout through the trigger interface units installed in every crate that participate 
in the readout process.

The trigger system is free-running and driven by the 250~MHz global clock and has essentially zero 
dead time at the occupancies expected for HPS. The trigger supervisor can apply dead time if 
necessary, for example on a `busy' or `full' condition from the front-end electronics. The system is 
designed to handle trigger rates above 50~kHz and has a latency set to $\approx 3~\mu$s to match 
that required by the SVT APV25 ASIC. During the Test run, for the most part the trigger system 
required only a single cluster in either the top or bottom Ecal halves and was tested to above 100~kHz 
trigger rates by lowering thresholds.

@@

In order to minimize cable length for the analog APV25 output signal the ATCA crate was located 
approximately 1~m from the beam line, next to our cable vacuum feed troughs.   
Before shielding with lead-blankets was arranged, we observed two failures of normally reliable ATCA 
crate power supplies, time-correlated to beam instabilities.

@@

\subsection{General Data Acquisition and Online Computing}
\label{sec:daq}
Every crate participating in the readout process contains a Readout Controller (ROC) that 
collects digitized information, processes it, and sends it on to the event builder. For the ECal, both 
VXS crates run ROC applications in a single blade Intel-based CPU module running under CentOS 
Linux OS. For the SVT DAQ, the ROC application runs on the external PC under RHEL. 
The event builder assembles information from the ROCs into a single event which is passed to the 
event recorder that writes it to a RAID5-based data storage system capable of handling up to 
100~MB/s. The event builder and other critical components run on multicore Intel-based multi-CPU 
servers. The DAQ network system is a network router providing 10Gbit/s high-speed connection 
to the JLab computing facility for long-term storage. For the Test run, both the SVT and ECal ROC had a 
1Gbit/s link to the network router.

@@

After a trigger is received, the amplitude  of every  APV25 analoge output  is sampled and digitized in 
six consecutive time bins, separated by roughly 25~ns.  
The typical pulse shape obtained is shown in Fig.~\ref{fig:pulse_shape}. As the figure demonstrates,  
the SVT was well timed-in to the trigger with the rise of the pulse at the 3rd sampling 
point.
\begin{figure}[]
\begin{center}
{\small
	\includegraphics[width=6cm]{figures/run1351_110513_samples_L1_top.pdf}
	\caption{Accumulation of six pedestal-subtracted samples from individual SVT channels associated 
	with hits on tracks.}
	\label{fig:pulse_shape}
}
\end{center}
\end{figure}
These hits are passed through a simple clustering algorithm which forms clusters by grouping adjacent 
strips with the position of a cluster on the sensor determined by the amplitude-weighted mean.
With a linear gain up to $\approx 3$~MIPs, the cluster charge for hits associated with a track follow 
the characteristic Landau shape. 
%, see Fig.~\ref{fig:cluster_pulse}.  
A noise level between $1.1-1.5\times 10^{3}$ electrons was established through multiple calibration 
runs giving a signal to noise ratio of $21-25$. Lab-based radioactive source tests was used to 
provide the absolute charge normalization.
%\begin{figure}[]
%\begin{center}
%{\small
%	%\includegraphics[width=7cm]{figures/run1351_mip_small.pdf}
%	\includegraphics[width=7cm]{figures/mip_top_layer_2.png}
%    	\caption{ The cluster charge distribution for hits associated with a track. }
%	\label{fig:cluster_pulse}
%}
%\end{center}
%\end{figure}

In order to find the time, $t_0$, and amplitude of each hit, the six samples from each channel are fitted 
to an ideal $CR-RC$ function. After clustering hits on a sensor, the hit time for each cluster is computed 
as the amplitude-weighted average of the individually fitted $t_0$ on each channel. The 
$t_0$-resolution is studied by comparing the cluster hit time with the average of all cluster hit times on 
the track, see Fig.~\ref{fig:tracktime}, which has the expected jitter due to clock phase and trigger, 
approximately 25~ns. 
\begin{figure}[]
\begin{center}
{\small
	\includegraphics[width=7cm]{figures/test_run_1351_hit_time_corrected_top_layer2_mod.pdf}
	\caption{
	The residual of individual cluster times with the average of all clusters on the track. }
	\label{fig:tracktime}
}
\end{center}
\end{figure}
After correcting for offsets from each sensor (time-of-flight, clock phase) and accounting for the 
correlation between the $t_0$ and track time,  the extracted $t_0$ resolution is 2.6~ns. This is somewhat 
worse than the approximately 2~ns resolution expected for S/N~25 which we attribute to the true 
pulse shape differing from our idealized fit function which will be improved in the future. Reducing the 
APV25 ASIC pulse shaping time will also improve time resolution. 
These results show that we can operate with the six sample readout mode of the APV25 chip and 
achieve time resolution adequate for pileup rejection during electron running in HPS.

@@

The integrated pulse of each FADC channel was converted to energy by  
subtracting a pedestal and apply a conversion factor to convert ADC counts to energy. 
The pedestals are measured using special runs where each trigger records 100 samples of signals from 
the APDs with 4~ns between each sample. The pedestals were extracted from the 
part of the window before the actual hit in the calorimeter. After subtracting the pedestal, modules with 
integrated pulse energy of about 73~MeV are clustered using a simple algorithm similar to the one 
deployed for the trigger (see Sec.~\ref{sec:trigger}). Due to the high effective readout threshold of 
$73$~MeV the average number of crystals in a cluster was $\sim 3$ and the simple clustering algorithm 
worked well for reconstruction of the detected shower energy. An average noise level of approximately 
15~MeV was measured in special pedestal runs.

@@

\subsection{Trigger Performance}
As described above the energy from each 
crystal is measured differently in the trigger and what is readout from the ECal. 
The trigger performance was studied by simulating the 
trigger for each event and comparing to how the events were actually triggered.
First, we simulate the FADC trigger path by converting from readout hits (with fixed-size window 
integration) to trigger hits (time-over-threshold integration). Secondly, the CTP clustering 
algorithm and the trigger decision from the SSP is simulated before we compare the trigger decision 
and trigger time to what was reported by the actual trigger. For every event, the trigger reports the trigger 
decision as a bit mask (top half, bottom half or both) and the time the trigger fired. To eliminate trigger 
bias, we use a tag and probe method: to study the trigger performance in one half of the ECal, we select 
events which triggered the other half and where there was exactly one probe cluster in the ECal half 
under study. We then measure trigger efficiency as the fraction of tagged events that fired the trigger in 
the probe half as a function of the probe cluster energy, shown in Fig.~\ref{fig:turnon}. 
\begin{figure}[ht]
\begin{center}
{\small
	\includegraphics[width=7cm]{figures/h_cl_E_probedata_eff_h_cl_E_probeMC_eff_dataMC_1351-v7-trig-tagBot.pdf}
	\caption{Trigger efficiency in both halves of the ECal for data and simulation as a 
	function of cluster energy.}
\label{fig:turnon}
}
\end{center}
\end{figure}
The trigger turn-on is slow and reaches an uneven plateau just below 1~GeV for two reasons;  
gain variations between different crystals lead to the threshold variations and the nonlinearity of 
the time-over-threshold integral means that the effective threshold is higher for clusters that span 
multiple crystals. The effective trigger threshold is therefore dependent on position and energy of 
the particle as well as cluster multiplicity.

As a cross-check, the turn-on from the trigger threshold was measured to be 1280 in units of ADC counts 
as expected. The threshold was not perfectly sharp because of uncertainties in the conversion from 
readout to trigger hits described above, but based on comparisons with simulation we found that the 
trigger worked exactly as specified.

>>>>

The trigger system is designed to select time coincidences of electromagnetic clusters in the 
top and bottom halves of the ECal. Figure~\ref{fig:hps_trigger_cal} shows a schematic overview 
of each stage of the system. 
 \begin{figure}[b]
\begin{center}
{\small
 \includegraphics[width=8cm]{figures/hps_trigger_cal}
\caption{Block diagram of the ECAL trigger system consisting of the FADC that samples and digitizes 
signals for each detector channel and sends them for cluster finding in the CTP. The CTP clusters are 
sent to the SSP where the final trigger decision is taken based on pairs of clusters in both halves of the 
ECal. The decision is sent to the Trigger Supervisor (TS) that generates the necessary signals to readout 
the sub-detectors.}
 \label{fig:hps_trigger_cal}
}
\end{center}
 \end{figure}
Each channel on the FADC board have an 
independent data path to send 5-bit pulse energy and 3-bit pulse arrival time information every 32~ns 
to a 
trigger processing board (CTP) in the same crate. Contrary to the readout path 
described in Sec.~\ref{sec:fadc}, this energy is a pedestal subtracted time-over-threshold sum with 
programmable offsets and minimum threshold discriminator for each channel. With input from all 
FADC channels, i.e. one half of the ECal, the CTP performs cluster finding and calculates cluster 
energy, shape and timing information. The 3x3 fixed-window, highly parallel, FPGA-based cluster 
algorithm simultaneously searches for up to 125 clusters with energy sum larger than 
the programmable energy threshold ($\approx270$~MeV). Crystals in the 
fixed-window are included in the sum if the leading edge of the pulse occurred within a 32~ns time 
window to take into account clock skew and jitter throughout the system.
%, the clustering algorithm can allow for crystal hits from up to seven previous or future 4~ns clock %cycles to contribute to the cluster sum.  
A separate sub-system board (SSP) receives the clusters (only the 
highest energy 3x3 window to deal with overlapping and very large clusters) from 
the top and bottom half CTP every 32~ns and searches for pairs of clusters in a 8~ns wide 
trigger window. The trigger supervisor (TS) generates all the necessary signals and controls the 
entire DAQ system readout through the trigger interface units installed in every crate that participate 
in the readout process.

The trigger system is free-running and driven by the 250~MHz global clock and has essentially zero 
dead time at the occupancies expected for HPS. The trigger supervisor can apply dead time if 
necessary, for example on a `busy' or `full' condition from the front-end electronics. The system is 
designed to handle trigger rates above 50~kHz and has a latency set to $\approx 3~\mu$s to match 
that required by the SVT APV25 ASIC. During the Test run, for the most part the trigger system 
required only a single cluster in either the top or bottom Ecal halves and was tested to trigger rates 
above 100~kHz by lowering thresholds.

@@

In order to minimize cable length for the analog APV25 output signal the ATCA crate was located 
approximately 1~m from the beam line, next to our cable vacuum feed-troughs.   
Before shielding with lead-blankets was arranged, we observed two failures of normally reliable ATCA 
crate power supplies, time-correlated to beam instabilities.

@@

\subsection{General Data Acquisition and Online Computing}
\label{sec:daq}
Every crate participating in the readout process contains a Readout Controller (ROC) that 
collects digitized information, processes it, and sends it on to the event builder. For the ECal, both 
VXS crates run ROC applications in a single blade Intel-based CPU module running CentOS 
Linux OS. For the SVT DAQ, the ROC application runs on the external PC under RHEL. 
The event builder assembles information from the ROCs into a single event which is passed to the 
event recorder that writes it to a RAID5-based data storage system capable of handling up to 
100~MB/s. The event builder and other critical components run on multicore Intel-based multi-CPU 
servers. The DAQ network system is a network router providing 10Gbit/s high-speed connection 
to the JLab computing facility for long-term storage. For the Test run, both the SVT and ECal ROC had a 
1Gbit/s link to the network router.

@@

After a trigger is received, the amplitude  of every  APV25 analoge output  is sampled and digitized in 
six consecutive time bins, separated by roughly 25~ns.  
The typical pulse shape obtained is shown in Fig.~\ref{fig:pulse_shape}. As the figure demonstrates,  
the SVT was well timed-in to the trigger with the rise of the pulse at the 3rd sampling 
point. Note that in the Test run the APV25 ASICs were operating with a 50~ns shaping time.  
\begin{figure}[]
\begin{center}
{\small
	\includegraphics[width=6cm]{figures/run1351_110513_samples_L1_top.pdf}
	\caption{Accumulation of six pedestal-subtracted samples from individual SVT channels associated 
	with hits on tracks.}
	\label{fig:pulse_shape}
}
\end{center}
\end{figure}
These hits are passed through a simple clustering algorithm which forms clusters by grouping adjacent 
strips with the position of a cluster on the sensor determined by the amplitude-weighted mean.
With a linear gain up to $\approx 3$~MIPs, the cluster charge for hits associated with a track follow 
the characteristic Landau shape. 
%, see Fig.~\ref{fig:cluster_pulse}.  
A noise level between $1.1-1.5\times 10^{3}$ electrons was established through multiple calibration 
runs giving a signal to noise ratio of $21-25$. Lab-based radioactive source tests was used to 
provide the absolute charge normalization.
%\begin{figure}[]
%\begin{center}
%{\small
%	%\includegraphics[width=7cm]{figures/run1351_mip_small.pdf}
%	\includegraphics[width=7cm]{figures/mip_top_layer_2.png}
%    	\caption{ The cluster charge distribution for hits associated with a track. }
%	\label{fig:cluster_pulse}
%}
%\end{center}
%\end{figure}

In order to find the time, $t_0$, and amplitude of each hit, the six samples from each channel are fitted 
to an ideal $CR-RC$ function. After clustering hits on a sensor, the hit time for each cluster is computed 
as the amplitude-weighted average of the individually fitted $t_0$ on each channel. The 
$t_0$-resolution is studied by comparing the cluster hit time with the average of all cluster hit times on 
the track, see Fig.~\ref{fig:tracktime}. 
\begin{figure}[]
\begin{center}
{\small
	\includegraphics[width=7cm]{figures/test_run_1351_hit_time_corrected_top_layer2_mod.pdf}
	\caption{
	The residual of individual cluster times with the average of all clusters on the track. }
	\label{fig:tracktime}
}
\end{center}
\end{figure}
After correcting for offsets from each sensor (time-of-flight, clock phase) and accounting for the 
correlation between the $t_0$ and track time,  the extracted $t_0$ resolution is 2.6~ns. This is somewhat 
worse than the approximately 2~ns resolution expected for S/N~25 which we attribute to the true 
pulse shape differing from our idealized fit function which will be improved in the future. Reducing the 
APV25 ASIC pulse shaping time will also improve time resolution. 
These results show that we can operate with the six sample readout mode of the APV25 chip and 
achieve time resolution adequate for pileup rejection during electron running in HPS.

@@

The integrated pulse of each FADC channel was converted to energy by  
subtracting a pedestal and apply a conversion factor to convert ADC counts to energy. 
The pedestals are measured using special runs where each trigger records 100 samples of signals 
from the APDs with 4~ns between each sample. The pedestals were extracted from the 
part of the window before the actual hit in the calorimeter. Modules with signal above the threshold  
are clustered using a simple algorithm similar to the one 
deployed for the trigger (see Sec.~\ref{sec:trigger}). Due to the high effective readout threshold of 
73~MeV the average number of crystals in a cluster was $\sim 3$ and the simple clustering 
algorithm worked well for reconstruction of the detected shower energy. An average noise level of 
approximately 15~MeV was measured in special pedestal runs.

@@

\subsection{Trigger Performance}
As described above the energy from each 
crystal is measured differently in the trigger and what is readout from the ECal. 
The trigger performance was studied by simulating the 
trigger for each event and comparing to how the events were actually triggered.
To eliminate trigger bias, we use a tag and probe method: to study the trigger performance in one half 
of the ECal, we select 
events which triggered the other half and where there was exactly one probe cluster in the ECal half 
under study. We then measure trigger efficiency as the fraction of tagged events that fired the trigger in 
the probe half as a function of the probe cluster energy, shown in Fig.~\ref{fig:turnon}. 
\begin{figure}[ht]
\begin{center}
{\small
	\includegraphics[width=7cm]{figures/h_cl_E_probedata_eff_h_cl_E_probeMC_eff_dataMC_1351-v7-trig-tagBot.pdf}
	\caption{Trigger efficiency in both halves of the ECal for data and simulation as a 
	function of cluster energy.}
\label{fig:turnon}
}
\end{center}
\end{figure}
The trigger turn-on is slow and reaches an uneven plateau just below 1~GeV for two reasons;  
gain variations between different crystals lead to the threshold variations and the nonlinearity of 
the time-over-threshold integral means that the effective threshold is higher for clusters that span 
multiple crystals. The effective trigger threshold is therefore dependent on position and energy of 
the particle as well as cluster multiplicity.

As a cross-check we simulate the FADC trigger path by converting from readout hits (with fixed-size 
window integration) to trigger hits (time-over-threshold integration). The CTP clustering 
algorithm and the trigger decision from the SSP is simulated before we compare the trigger decision 
and trigger time to what was reported by the actual trigger. For every event, the trigger reports the 
trigger decision as a bit mask (top half, bottom half or both) and the time the trigger fired.
The turn-on from the trigger threshold was measured to be 1280 in units of ADC counts as expected. 
The threshold was not perfectly sharp because of uncertainties in the conversion from readout to trigger 
hits described above, but based on comparisons with simulation we found that the 
trigger worked exactly as specified.
